---
title: "Momocs speed dating"
author: "Vincent Bonhomme"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
#output: 
#  html_document:
#    toc: true
#    toc_depth: 3
#    highlight: tango
    
vignette: >
  %\VignetteIndexEntry{Momocs speed dating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(eval = TRUE, message=FALSE, 
               warnings=FALSE, results="hold")
```

## Preliminaries
### Abstract
Momocs aims to provide a complete and convenient toolkit for morphometrics. It is intended for scientists interested in describing quantitatively the shape, and its variations, of the objects they study.

In the last decade, R has become the open-source lingua franca for statistics, and morphometrics known its so-called 'revolution'. Nevertheless, morphometric analyses still have to be carried out using various software packages, for which source code is mostly unavailable or copyrighted. Moreover, most of existing software packages cannot be extended and their bugs are hard to detect and thus correct. This situation is detrimental to morphometrics; time is wasted, analyses are restricted to available methods, and last but not least, are poorly reproducible. This impedes collaborative effort both in software development and in morphometric studies.

By gathering the common morphometric approaches in an open-source environment and welcoming contributions, Momocs is an (work-in-progress) attempt to solve this twofold problem and to push morphometrics one step further.

It hinges on the core functions published in the must-have _[Morphometrics with R](http://www.springer.com/statistics/life+sciences,+medicine+%26+health/book/978-0-387-77789-4)_ by [Julien Claude](http://www.isem.univ-montp2.fr/recherche/equipes/biologie-du-developpement-et-evolution/personnel/claude-julien/) (2008), but has been further extended to allow other shape description systems. So far, configurations of landmarks, outlines and open outline analyses, along with some facilities for traditional morphometrics have been implemented.

Prior to analysis, Momocs can be used to acquire and manipulate data or to import/export from/to other formats. Momocs also has the facility for a wide range of multivariate analyses and production of the companion graphics. Thus a researcher will find that just a few lines of code will provide initial results, but the methods implemented can be finely tuned and extended according to the user's needs.

* See Momocs [features](https://github.com/vbonhomme/Momocs/#features).
* If you use it, cite it: `citation("Momocs")`.
* This citation refers to an obsolete version of Momocs, only handling outline analyses. The next one is in preparation.

### Survival tips
* This vignette gently introduces Momocs; more specific help can be find in function's helfiles, for instance `?efourier`.
* There is an online (and much nicer) version of this manual that can be accessed with `Momocs_help("efourier")`.
* Feel free to contribute to Momocs through [GitHub](http://www.github.com/vbonhomme/Momocs): report issues, ask for new features, share data and methods, correct typos, write better vignettes, helpfiles, or whatever pleases you. If you have never heard of GitHub, that's definitely worth [a look](http://r-pkgs.had.co.nz/git.html).
* Feel free to drop me a line, should you need a hand or would like to collaborate with me:   `bonhomme.vincent@gmail.com`.

### Get, install and use it
First, of all, let's download the last version of Momocs. You will need to install the `devtools` package to get it from my [GitHub](http://www.github.com/vbonhomme) repository :

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("vbonhomme/Momocs")
```

The typical `install_packages("Momocs")` is not recommended as Momocs is still under active development, not compatible with frequent, time-consuming, and overall boring CRAN submission.

We can start using Momocs, as long as it has been loaded with :

```{r, eval=TRUE, echo=TRUE, message=FALSE}
library(Momocs)
```

## Design
Keywords used all accross Momocs are introduced here in __bold__.

Morphometrics is the ugly job of turning beautiful shapes into quantitative variables. Just kidding, that's pretty exciting.

A __shape__ is defined as a collection of `(x; y)` coordinates. No 3D yet but different __families__ can be handled: __outlines__, here in a first-quarter moon ; __open outlines__, here is the sterile valve of an olive stone; __configuration of landmarks__; here, hologous points from a mosquito wing.

```{r, echo=FALSE, message=FALSE}
shapes[18] %>% coo_sample(60) %>% coo_plot(points=TRUE)
olea[11] %>% coo_sample(32) %>% coo_plot(points=TRUE)
wings[1] %>% coo_plot(points=TRUE)
```

They are all single shapes defined by a matrix of `(x; y)` coordinates; here are the first points of the moon:

```{r}
shapes[18] %>% head()
```
A few dozens of operations on single shapes are implemented such as: plotting, centering, calculating areas, etc. They can be accessed with `apropos("coo_")`. But working on single shapes is quite boring.

Shapes can be organized into a __collection__ of coordinates: a `Coo` object that carries:

* always, a component named `$coo`, a `list` of shapes (as `matrix`.ces);
* most of the time, a component named `$fac`, a `data.frame`;
* possibly, other components of interest (thier inventory can be accessed with `ls()`).

One can do several things with a `Coo` object: visualize it, apply morphometric operations, handle the data it contains, but in the end, a __ morphometric method__ to turn coordinates into coefficients.

Morphometrics on coordinates produce a collection __collections__ of coefficients: a `Coe` objects that carries:

* always, a component named `$coe`, a `matrix` of coefficients;
* if present in `Coo`, `$fac` is copied;
* possibly, other components of insterest (their inventory can be accessed with `ls()`).

This can be summarized as follows:

`Coo`                      | + | Morphometric method  | = | `Coe`
---------------------------|----|----------------------|----|---------------
`(x;y)` coordinates        | +  | appropriate method   | =  |quantitative variables


Thus `Coo` objects are collections of coordinates that become `Coe` objects when an appropriate morphometric method is applied on them.

The different families of shape sometimes implies or requires particular operations on them. Some operations on shapes are __generic__ in that they do not depend of the family. For instance, centering a configuration of landmarks or an outline, or calculating their centroid size is, mathematically, the same __generic__ operation. But some operations on shapes are __specific__ to a peculiar family. For instance, calculating elliptical Fourier transforms on a configuration of landmarks makes no sense. 

Momocs implement this desirable behavior and defines __classes__ and __subclasses__, as S3 objects.

`Coo`                                | Morphometrics methods              | `Coe`  
-------------------------------------|------------------------------------|----------  
`OutCoo` (outlines)                  | `efourier`, `rfourier`, `tfourier` | `OutCoe`
`OpnCoo` (open outlines)             | `npoly`, `opoly`, `dfourier`       | `OpnCoe`  
`LdkCoo` (configuration of landmarks)|`fgProcrustes`, `slide`             | `LdkCoe`

In other words:

* any collection of shapes belong to: primarily (pick one) `{OutCoo, OpnCoo, LdkCoo}`; and `Coo` classes
* generic and specific methods (typically morphometrics methods) can be applied to it
* it results in a collection of coefficients that belong to: primarily (pick one) `{OutCoe, OpnCoe, LdkCoe}`; and `Coe` classes

Finally, generic and specific operations can be applied to the `Coe` objects, chiefly multivariate methods, capitalicized: `PCA`, `LDA`, `CLUST`, `MANOVA`, etc.

Overall, Momocs implements a simple and consistent grammar that is detailed below. 

Last but not least, if you're familiar with modern R and the Hadley-verse, you should feel home as `ggplot2` graphics, `dplyr` verbs and `magrittr` pipes are implemented available.

## Single shapes
Let's load one of the Momocs datasets, some various outlines (an `Out` object):

```{r}
shapes                    # prints a brief summary
panel(shapes, names=TRUE) # base graphics
panel2(shapes)            # ggplot2 graphics
```

`shapes` is one of the datasets bundled with Momocs. It's ("lazy") loaded in memory as soon as you call it, no need for `data(shapes)`. To see all Momocs' datasets, try `data(package="Momocs")`. These datasets are all `Coo` objects (try `class(bot)`), ie collection of shapes. One can do many things on a `Coo` object, as above, eg printing a summary of it (just by typing its name), plotting a family picture with `panel` or `panel2`. Note the `2` that refers to a `ggplot2` variant of a given plot.

So far, we're interested in single shapes so 
let's extract the 4th shape from `shapes`, using the traditional syntax.  We plot it with `coo_plot` that comes with several options for plotting all families of shapes.

```{r}
shp <- shapes[4]
coo_plot(shp)
# coo_plot is the base plotter for shapes
# but it can be finely customized, see ?coo_plot
coo_plot(shp, col="grey80", border=NA, centroid=FALSE, main="Meow")
```

Let's now do some basic operations on this shape. They all named `coo_*` and you can have the full list with `apropos("coo_")`. `coo_*` family encompasses:

 * geometric operations (such as centering, scaling, etc.)
 * plotting functions 
 * scalar descriptors of shape (such as area, perimeter, circularity, rectilinearity, etc.)
 * various other operations on a single shape.

```{r}
coo_plot(coo_center(shp), main="centered Meow")
coo_plot(coo_sample(shp, 64), points=TRUE, pch=20, main="64-pts Meow")
```

Momocs makes use of [maggritr](http://cran.r-project.org/package=magrittr)'s pipe operators. A nice introduction can be found [there](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html). `magrittr` requires a (very small) cerebral gymnastics at the beginning but the benefits are huge, for R as a whole. It makes things clearer, it: saves typing; reduces intermediate variable assignation; reads from left to right; substantiates the pipe we (should) have in mind. `magrittr`'s pipes are already loaded with Momocs.

```{r}
shapes[4] %>% coo_smooth(5) %>% coo_sample(64) %>% coo_scale() %>% coo_plot()

# pipes can be turned into custom function
cs64 <- function(x) x %>% coo_sample(64) %>% coo_scale() %>% coo_center()
shapes[4] %>% cs64 %>% coo_plot() # note the axes
```

The most familiar operation can directly be applied on `Coo` objects:
```{r}
bot %>% 
  coo_center %>% coo_scale %>% 
  coo_alignxax() %>% coo_slidedirection("N") %T>% 
  print() %>% stack()
```
Notice that before the `stack` operation, we still have a `Coo` object. And by the way, pipes can also be defined on Momocs objects.

## Morphometrics
### Outline analysis
A word about data import: you can extract outlines from a list of black masks over a white background, as `.jpg` images with `import_jpg`. Have a look to helpfiles (`import_jpg` and `import_jpg1`) for more details. Here we do not bother with import since we will use the `bot`tles outlines dataset bundled with Momocs.

```{r}
data(bot)
bot
panel(bot, fac="type", names=TRUE)
stack(bot)
````

Here we will illustrate outline analysis with some elliptical Fourier transforms (but the less used - and tested - radii variation Fourier transforms and tangent angle Fourier transforms are also implemented with `rfourier` and `tfourier`).

The idea behind elliptical Fourier transforms is to fit the x and y coordinates separately, that is the blue and red curves below:
```{r}
coo_oscillo(bot[1], "efourier")
```

Graphically, this is equivalent to fitting Ptolemaic ellipses on the plane:

```{r}
Ptolemy(bot[1])
```

Let's calibrate the number of harmonics required. More details can be found in their respective help files.

```{r}
calibrate_harmonicpower(bot)
calibrate_deviations(bot)
calibrate_reconstructions(bot)
````

Here, 10 harmonics gather 99% of the harmonic power. If you're happy with this criterium, you can even omit `nb.h` in `efourier` : that's the default parameter.

```{r}
bot.f <- efourier(bot, nb.h=10)
bot.f
```

`bot.f` is a `Coe` object (and even an `OutCoe`), you have have a look to the help files to go deeper into Momocs classes.

Let's have a look to the amplitude of fitted coefficients.
```{r}
hist(bot.f, drop=0)
boxplot(bot.f, drop=1)
```

Now, we can calculate a PCA on the `Coe` object and plot it, along with morphospaces, calculated on the fly.

```{r}
bot.p <- PCA(bot.f)
class(bot.p)        # a PCA object, let's plot it
plot(bot.p)
```

Amazing but we will do much better afterwards.

The question of normalization in elliptical Fourier transforms is central. Normalization can either be done beforehand, with geometric operations, or afterhand, directly on the matrix of Fourier coefficients and consuming the first harmonic. In brief, I'm not a big fan of the "us the first harmonics and see what happens" strategy, as some biases can be introduced (and actually quite hard to detect), particularly on rounded/ellipsoid shapes. More can be found in `?efourier`.

Here is an example of such bias from the `molars` dataset generously shared by Cornu and Detroit.

```{r, message=FALSE, error=FALSE, warning=FALSE, results="hide"}
# raw molars dataset
stack(molars, title = "Non-aligned molars")
# Procrustes-aligned and slided molars
mol.al <- fgProcrustes(molars, tol = 1e-4) %>% coo_slidedirection("W")
stack(mol.al, title="Aligned molars")

# Now compare PCA and morphospace using the 1st harmonic alignment
molars %>% efourier(norm=TRUE) %>% PCA() %>% plot("type")
# and the a priori normalization 
molars %>% efourier(norm=FALSE) %>% PCA() %>% plot("type")
```

Finally, you can drop some harmonics with `rm_harm`. And methods that removes the bilateral (a)symmetry are implemented: `rm_Asym` and `rm_Sym`, while `symmetry` calculates some related indices.

### Open outlines
Open outlines are curves. Methods actually implemented are:

* `npoly` that fit natural polynomials;
* `opoly` that fit orthogonal (also called Legendre's) polynomials;
* `dfourier` for the discrete cosine transform.

Note that `opoly` and `npoly` can only be used on _simple_ curves, curves that have at most one `y` for any `x` coordinates, at least under a given orientation. `dfourier` can fit _complex_ curves, curves "that bak on their feets".

Here, we will work on the fertile valves of olive stones, a (very partial) dataset provided by my colleagues Terral, Ivorra, and others.

They have two orthogonal views (a lateral and a dorsal view). See the paper cited in `?olea` for more details. Let's explore it a bit:

```{r}
olea
stack(olea, fac="view")     # already aligned \o/
panel(olea, names="ind")    # another family picture
```

Now, we gonna calculate `opoly` on it and plot the result of the PCA. Notice how consistent is the grammar and the objects obtained:

```{r}
op <- opoly(olea)           # orthogonal polynomials
class(op)                   # an OpnCoe, but also a Coe
op.p <- PCA(op)             # we calculate a PCA on it
class(op.p)                 # a PCA object
plot(PCA(op), ~domes+var)   # notice the formula interface to combine factors
```

But this is perfectly wrong! We merged the two views are if they were different individuals. Momocs can first `chop` or `filter` the whole dataset to separate the two views, do morphometrics on them, and `combine` them afterwards.

```{r, message=FALSE}
table(olea, "view", "var") 
# we drop 'Cypre' since there is no VL for 'Cypre' var
olea %>% filter(var != "Cypre") %>%              
  # split, do morphometrics, combine
  chop(view) %>% lapply(opoly) %>% combine() %T>%
   # we print the OpnCoe object, then resume to the pipe
  print() %>%
  # note the two views in the morphospace
  PCA() %>% plot("var")
```

Now the PCA is done on the combination of two `OpnCoe` objects, each one resulting from an independant `opoly` call. That is the meaning of the `[ combined: opoly + opoly analyses ]` printed by the pipe above. Momocs can combine up to four different views.

### Configuration of landmarks
_Landmarks methods are being implemented in Momocs (as of nov 2015). Full generalized Procrustes methods are available, and sliding methods on semi landmarls will soon be. Here is a first glimpse._

First, let's have a look to graphics facilities
```{r, message=FALSE, echo=FALSE}
stack(wings)
w.al <- fgProcrustes(wings)
stack(w.al)
ldk_confell(w.al$coo, col = "red")
ldk_chull(w.al$coo)
ldk_contour(w.al$coo)
ldk_labels(mshapes(w.al$coo))
```

### Traditionnal morphometrics
Traditional morphometrics lose geometries: from the variables, you can't unambiguously reconstruct the shape. Every shape is described by a combination of measurements, (inter landmark distance, quantitative variables, scalar descriptor, etc.)

Momocs provides some basics utilities to work with such objects in the `TraCoe` class. There is not `TraCoo` per se but it can be obtained from any `Coo` with the `measure` method. Let's take the `hearts` dataset that comes from handdrawn heart shapes from my former colleagues at the French Intitute of Pondicherry:

```{r}
hearts
panel(hearts, fac="aut", names="aut")
```

Notice that there are 4 landmarks defined on them. Such landmarks on outlines can be: defined with` def_ldk()`, retrieved with `get_ldk()`, and overall used to align outlines with `fgProcrustes()`. You can compare: `hearts %>% stack()` with `hearts %>% fgProcrustes() %>% coo_slide(ldk=1) %>% stack()`.

Let's describe these hearts with scalar descriptors: area, circularity and the distance between the 1st and the 3rd bumps of the hearts. `measure` is of great help. Note the loadings.

```{r}
ht <- measure(hearts, coo_area, coo_circularity, d(1, 3))
ht %>% PCA() %>% plot("aut", pch=20, ellipsesax=F, ellipse=T, loadings=T)
```

Again, there are plenty of scalar descriptors of shape, which names starts with `coo_*`, `apropos("coo_")`.

Such a `TraCoe` is provided in the `flower` dataset which is simply a rearranged `iris`. Once again, note the grammar consistency.

```{r}
flower
flower %>% PCA() %>% plot("sp", loadings=TRUE)
```

Note that, by default, PCA on `TraCoe` object first centers and scales variables. This can be changed, see `?PCA`.


## Multivariate statistics

This section will mainly be illustrated with `bot`, and consequently outline analysis, but it works the same on any `Coe` object, resulting from open outlines, configuration of landmarks, traditional morphometrics, etc.

```{r, message=FALSE}
bot.f <- efourier(bot)
```


### `PCA`: Principal Component Analysis
Let's see the main components of shape variability with a Principal Component Analysis.

```{r}
bot.p <- PCA(bot.f)
plot(bot.p)
````

Morphological spaces are reconstructed on the fly with `plot.PCA`. We call it `plot.PCA` because it uses the familiar `plot` but on the particular `PCA` class (type `class(bot.p)`). We may want to display the two groups saved in `bot$fac`. Just type the id of the column or its name. 
```{r}
plot(bot.p, "type") # equivalent to plot(bot.p, 1)
```

`plot.PCA` has many arguments and examples below will give a glimpse, but see `?plot.PCA` for the full list.

```{r}
plot(bot.p, 1, ellipses=TRUE, ellipsesax = FALSE, pch=c(4, 5))
plot(bot.p, 1, chull=TRUE, pos.shp = "full_axes", abbreviate.labelsgroups = TRUE, points=FALSE, labelspoints = TRUE)
plot(bot.p, 1, pos.shp="circle", stars=TRUE, palette=col_qual)
```

You have a `ggplot2` alternative, that is not completely consistent:
```{r}
plot2(bot.p, "type")
```

And other helper functions for the `PCA` class:
```{r}
scree(bot.p)
scree_plot(bot.p)
boxplot(bot.p, 1)
PCcontrib(bot.p)
```

By the way, you can use Momocs plotters to plot non-morphometric datasets. Using a `TraCoe` object is an option, but `PCA` also works fine. Let's see an example with `iris` dataset:

```{r}
TraCoe(iris[, -5], fac=data.frame(sp=iris$Species)) %>% 
  PCA() %>% plot("sp", loadings=TRUE)
# or
PCA(iris[, -5], fac=data.frame(sp=iris$Species)) %>%
  plot("sp", chull=TRUE, ellipses=TRUE, conf_ellipses = 0.9)
```

### `LDA`: Linear Discriminant Analysis
We can also calculate a Linear Discriminant Analysis on the PCA scores, or on the `Coe` object, directly on the matrix of coefficients but we may encounter collinearity between variables. Try the following:
```{r, eval=FALSE}
#LDA(bot.f, 1)
# we work on PCA scores
bot.l <- LDA(bot.p, 1)
# print a summary, along with the leave-one-out cross-validation table. 
bot.l 
# plot.LDA works pretty much with the same grammar as plot.PCA
# here we only have one LD
plot(bot.l)
# plot the cross-validation table
plot_CV(bot.l)  # tabular version
plot_CV2(bot.l) # arrays version
```

Finally, if you need to classify using LDA methods, see `classify`.

### `MANOVA`: Multivariate Analysis of variace
We can test for a difference in the distribution of PC scores with:
```{r}
MANOVA(bot.p, "type")
```

We can also calculate pairwise combination between every levels of a `fac`. Here we just have two levels, so a single pairwise combination but the syntax is:

```{r}
MANOVA_PW(bot.p, "type")
```

### `CLUST`: Hierarchical clustering
A hierarchical classification now. It relies on `ggtree` which is itself under active development. So you may have trouble to make this command work.
```{r, eval=FALSE}
CLUST(bot.p, 1)
```
Monophyly is plotted by default. Many options can be found in `?CLUST`

### `KMEANS`: K-means clustering
A very minimal k-means clustering is implemented:
```{r}
KMEANS(bot.p, centers = 5)
```

### `mshapes`: Mean shapes
We can retrieve the mean shapes, group wise (if a `fac` is specified), or the global mean shape (if omitted). It works from the `Coe` object:
```{r}
# mean shape
bot.f %>% mshapes() %>% coo_plot()
# mean shape, per group
bot.ms <- mshapes(bot.f, 1)
beer   <- bot.ms$shp$beer    %T>% coo_plot(border="blue")
whisky <- bot.ms$shp$whisky  %T>% coo_draw(border="red")
legend("topright", lwd=1,
       col=c("blue", "red"), legend=c("beer", "whisky"))
```

## Manipulating objects
### `dplyr` verbs
One common yet boring task of morphometrics consists in handling datasets: add new information, remove some individuals, etc.

Momocs adapts [dplyr](http://cran.r-project.org/package=dplyr) verbs to its objects, and add new ones. If you have never heard of dplyr, let's have a look to this [introduction](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html), this may change your (R) life.

Basics verbs are implemented; try the following:
```{r, results="markup"}
data(olea)
olea

# slice: select individuals based on their position
slice(olea, 1:5)
slice(olea, -(1:100))

# filter: select individual based on a logical condition
filter(olea, domes=="cult", view!="VD")

# select: pick, reorder columns from the $fac
select(olea, 1, Ind=ind)

# rename: rename columns (select can also do it)
rename(olea, domesticated=domes)

# mutate: add new columns
mutate(olea, fake=factor(rep(letters[1:2], each=105)))

# transmute: add new columns and drop others
transmute(olea, fake=factor(rep(letters[1:2], each=105)))
```

And you can pipe those operations: say, we only want dorsal views from domesticated individuals, for a (renamed) 'status' column, and drop the 'ind' column:
```{r}
olea %>% 
filter(domes=="cult", view=="VD") %>% 
rename(domesticated=domes) %>% 
select(-ind)
```

This should save some headaches.

### New verbs ala `dplyr`
New verbs are implemented: for instance, you can `chop` (a rougher slicing) according to a condition: this will create a list, on which you can apply further operations, then `combine` it back. This is particularly useful when you want to apply independant treatments to different partitions, eg orthogonal views of your model. Prior to this, we can use `table` to cross-tabulate data from `$fac`. We could have done the first step of what follows with `rm_uncomplete` that drops (if any) missing data.

```{r}
table(olea, "view", "var") 
# we drop 'Cypre' since there is no VL for 'Cypre' var
olea %>% filter(var != "Cypre") %>%              
  # split, do morphometrics, combine
  chop(view) %>% lapply(opoly) %>% combine() %>% 
  # note the two views in the morphospace
  PCA() %>% plot("var")
```

### Various helpers
Some methods help, on `Coe` objects to:

 * sample a given number: `sample_n`;
 * sample a given proportion: `sample_frac`;
 * generate new individuals based on calibrated Gaussian coefficient generation: `breed`;
 * generate new individuals based on permutations: `perm`.

Several shortcuts are implemented on `Coo` and `Coe` objects:
* `names` returns shape names;
* `length` returns their number;
* `table` cross-tabulates the`$fac` component;
* `Ntable` does the same job and plots a confusion matrix;
* `[]` extracts one (or more) shape;
* `$` can access either a shape name or a column name for the `$fac`.

Try the following:
```{r, eval=FALSE}
names(bot)
length(bot)
table(olea, "var", "domes")
Ntable(olea, "var", "domes")
bot[1]
bot[1:5]
bot$brahma
bot$type
```

## Babel import/export
There are various morphometrics formats in the wild, almost as much as softwares. Momocs tries to create bridges between them, all gathered in the **Babel** family.

### Bridges within R
You can convert from/to array, matrix, list or data.frame with the functions `{a, m, l, d}2{a, m, l, d}`. For instance, `l2a` converts a list into an array that you can use with `geomorph`; `a2l` does the inverse operation.

Imagine you want to import pupfish from `geomorph` as a `Ldk` object:

```{r, eval=FALSE}
library(geomorph)
data(pupfish)
str(pupfish)
# so $coords will become $coo, and
# all other components will be turned into a data.frame to feed $fac
# with a single line
Ldk(coo=pupfish$coords %>% a2l,
fac=pupfish[-1] %>% as.data.frame())
```

### Import from StereoMorph
If you use this excellent package by Olsen to digitize landmarks and curves, you can import them, from the files produced with the functions `import_StereoMorph_ldk` and `import_StereoMorph_curve`.

### Import from tps and other digitizing softwares
* `.tps` files can be read with `import_tps`
* `.nts` files an be read with `nts2Coo` (will be turned into `import_nts` soon)

### Import misc
* If covariables are encoded in filenames, which is a good practice, eg if you have files named `spA_group7_ind4_VL.{txt|jpg|etc.}`, use `lf_structure`;
* If you need rewriting rules on your `$fac`, `rw_rule` is your friend;
* If you need to rescale imported coordinates, see `rescale`;
* If you need to tie images (eg outlines) and `.txt` (eg coordinates of landmarks on them), see `tie_jpg_txt`.

### Save from R
The best way to save a Momocs object is probably to use the base `save` function. You can call it back afterwards with `load`:

```{r, eval=FALSE}
save(bot, file="Bottles.rda")
# closing R, going to the beach, back at work
load("Bottles.rda") # bot is back
```

### Export from R
Any Momocs object, `Coo`s, `Coe`s, `PCA`s, etc. can be turned into a `data.frame` with `as_df`. Work with `dplyr`, `ggplot2` is made easy and you can export it as `.txt`, `.csv` "by hand" or use the `export` function:

```{r, eval=FALSE}
bot %>% as_df # then %>% write.table
bot %>% efourier %>% export
bot %>% efourier %>% PCA %>% export
```

But, of course, you can directly access information within the Momocs objects; try the following:

```{r, eval=FALSE}
# from Coo objects
bot$coo
bot$fac

# from Coe objects
bot.f$coe
bot.f$fac
as_df(bot.f)

# from PCA objects
bot.p$x # scores
bot.p$rotation # rotation matrix
```

## Graphics
The most common graphics have been already presented in previous sections. But there are many more of them, and they are gathered in the 
**Graphics** family that can be accessed with ... TODO.

### `tps_*`: Thin Plate Splines
TPS have not been presented before but here there are:
```
tps_grid(beer, whisky)
tps_arr(beer, whisky)
tps_iso(beer, whisky)
```

Again, plenty options in `?tps_*`.


### `ggplot2`
Many graphical function have their `ggplot2` counterpart:

```{r}
panel(bot)
panel2(bot)
```

And the classical `ggplot2` grammar applies:

```{r}
library(ggplot2)
gg <- panel2(bot)
gg + theme_minimal()
```

Otherwise, you can build your own plots. Let's begin with a simple example:
```{r}
# we build a ggplot object from a shape turned into a data.frame
shapes[4] %>% m2d() %>% ggplot() + 
  aes(x, y) + geom_path() + coord_equal() + 
  labs(title="ggplot2 Meow") + theme_minimal()
```

Here is a more complicated plot that takes profit of `as_df` methods that turns objects into `ggplot2`-friendly `data.frame`s:

```{r}
bot.p %>% as_df() %>% ggplot() +
  aes(x=PC1, y=PC2, col=type) + coord_equal() + 
  geom_point() + geom_density2d() + theme_light()
```

### Color palettes
 * Nice palettes are bundled with Momocs, see `?Palettes`.
 * Overall, [RColorBrewer](http://cran.r-project.org/package=RColorBrewer) is more than an option if you work with colors.

## Frequently asked questions

  * __How to access all methods for a given class?__
    + `methods(class="your_class")`, eg `methods(class="Opn")`
    
  * __How to access methods defined for a particular function?__
    + `methods("your_function")`, eg `methods("fgProcrustes")`
    
  * __How can I access help for a particular methods?__
    + `function.class`, eg `?plot.PCA`
    
  * __How can I access the code of a method?__
    + For functions, just type its name, eg `lf_structure`; for methods add its class, eg `plot.PCA`
  
  * __To be continued__
  
------
 * Done with love by Vincent Bonhomme `<bonhomme.vincent@gmail.com>`.
 * Sources are [there](https://github.com/vbonhomme/Momocs/tree/master/vignettes)
