---
title: "Operations on shapes"
author: "Vincent Bonhomme"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{Momocs_coo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preamble
In Momocs, a shape is a matrix of `(x, y)` coordinates. When they are plenty, they are usually gathered in a `Coo` object which is, essentially, a list with at least a `$coo` component.

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(Momocs)
```

```{r}
# a single shape
bot[1] %>% is_shp() # tests if it is a 2-col matrix without NAs
bot[1] %>% head()

# a collection of shapes
bot %>% class()     # Out for outlines, but also a Coo. See ?Coo 
bot %>% is.list()   # essentially a list
bot$coo %>% class() # with a $coo component
bot$coo[1:2] %>% lapply(head) # where shapes are gathered
```

Most operations on shapes can also be done on `Coo` objects. For example, getting the area or centering:

```{r}
coo_area(bot[1]) # in pixels^2
coo_area(bot) %>% head()
```
In other words most `coo_` functions are actually _methods_ that can actually be passed with a single matrix or a Coo object.

Below, we will distinguish `coo_*` operations based on whether they return a shape (eg if it is a geometric operation) or a scalar (a single number that can be used as a shape descriptor).

Not all are described but you can obtain a full list of `coo_` operations with:
```{r}
cool <- apropos("coo_") # get the full list
head(cool)              # the firt 6
length(cool)            # how many are defined
```



## Scalar descriptors of shape
`coo_scalar` calculate _all_ scalar descriptors included in Momocs. `coo_rectilinearity` is not included by default since it takes a looot of time to compute.

```{r}
df <- shapes %>% coo_scalars()
df
```

Each of the function listed below has its own method that you can used by prefixing it with `coo_`; see the example below with `coo_area`.
```{r}
colnames(df) %>% cat(sep="\n")

coo_area(shapes[4]) # in pixels^2
coo_area(shapes)

# a subset
dplyr::data_frame(area=coo_area(shapes),
                  perim=coo_perim(shapes)) %>% head()
```

## Geometric operations
The functions below return shapes, so we exemplify them with graphs. For the sake of clarity/speed we illustrate them on single shape but, again, this can be done on `Coo` objects. 
```{r}
# A single example on Coo
bot %>% stack
bot %>% coo_center %>% stack()
```

Now we go with a cat and we will extensively use `coo_plot`. We define a function that will help display side by side the original and transformed cat; when I use `p(coo_align)`, it is equivalent to `coo_align(your_shp)` or `shp %>% coo_align`
```{r}
shp <- shapes[4]
shp %>% coo_plot()
```
```{r, include=FALSE}
# from a shape (shp) in the global environment (do not tell anyone !)
# apply a transformation (fun) and plot the two, side by side
p <- function(fun){
  title <- as.character(match.call()$fun) # capture the function name
  par(mfrow=c(1, 2))                      # side by side plots
  shp %>% coo_plot()                      # original shape
  shp %>% fun %>% coo_plot(main=title)    # transform and plot
  par(mfrow=c(1, 1))                      # back to original layout
}
```

### Size
#### Isotropic
```{r}
shp %>% coo_plot(main="original cat")
# scaling
shp %>% coo_scale %>% coo_plot(main="coo_scale")
# templating 
shp %>% coo_template(1) %>% coo_plot(ylim=c(-1, 1), main="coo_template"); rect(-0.5, -0.5, 0.5, 0.5)

```

#### Anisotropic
```{r}
shp %>% coo_scalex(2) %>% coo_plot(main="coo_scalex")
shp %>% coo_scaley(3) %>% coo_plot(main="coo_scaley")
shp %>% coo_shearx(1) %>% coo_plot(main="coo_shearx")
shp %>% coo_sheary(1) %>% coo_plot(main="coo_sheary")
```


### Alignment
#### Based on geometry
```{r}
shp %>% coo_align          %>% coo_plot(main="coo_align")
shp %>% coo_aligncalliper  %>% coo_plot(main="coo_aligncalliper")
shp %>% coo_alignminradius %>% coo_plot(main="coo_alignminradius")
shp %>% coo_alignxax       %>% coo_plot(main="coo_alignxax")
```

#### Based on landmarks

```{r}
# we use points with id 93 and 516 as landmarks
coo_bookstein(shp, 93, 516) %>% coo_plot(main="coo_bookstein")
# see coo_baseline for a more generic version
```
### Translation
```{r}
shp %>% coo_center %>% coo_plot(main="coo_center") # or coo_centre if you come from UK
shp %>% coo_center %>% coo_trans(x=  5) %>% coo_plot(main="coo_trans")
shp %>% coo_center %>% coo_trans(y= -2) %>% coo_plot(main="coo_trans")
```

### Rotation
```{r}
shp %>% coo_rotate(pi/2) %>% coo_plot(main="coo_rotate(pi/2)")
shp %>% coo_rotatecenter(-pi/6, center=c(250, 195)) %>% coo_plot(main="coo_rotatecenter") # with center ~on cat's nose
# mirrorring
shp %>% coo_flipx %>% coo_plot(main="coo_flipx")
shp %>% coo_flipy %>% coo_plot(main="coo_flipy")
```

### Sampling
```{r}
shp %>% coo_nb()
shp %>% coo_sample(36) %>% coo_plot(main="coo_sample")
shp %>% coo_sample_prop(1/20) %>%  coo_plot(main="coo_sample")
shp %>% coo_samplerr(36) %>% coo_plot(main="coo_samplerr") # regular radius
shp %>% coo_sample(36) %>% coo_interpolate(144) %>% coo_plot(points=TRUE, main="coo_interpolate")
```

### Smoothing
```{r}
shp %>% coo_sample(24) %>% coo_plot(main="a cubist cat")
shp %>% coo_sample(24) %>% coo_smooth(2) %>% coo_plot(main="coo_smooth")
```

If you have curves, you may prefer not to change first and last points:
```{r}
olea[1] %>% coo_plot()
olea[1] %>% coo_smooth(12) %>% coo_plot(main="coo_smooth")
olea[1] %>% coo_smoothcurve(12) %>% coo_plot(main="coo_smoothcurve")
```


### Slicing
#### Based on geometry
```{r}
shp %>% coo_center %>% coo_plot(main="centered cat")
shp %>% coo_center %>% coo_up() %>% coo_plot(main="coo_up")
shp %>% coo_center %>% coo_down() %>% coo_plot(main="coo_down")
shp %>% coo_center %>% coo_left() %>% coo_plot(main="coo_left")
shp %>% coo_center %>% coo_right() %>% coo_plot(main="coo_right")
```
If you want to slice but want the first point to be on one end, then `coo_slidegap` is your friend :
```{r}
shp %>% coo_center %>% coo_right() %>% coo_slidegap %>% coo_plot(main="coo_right")
```

#### Based on landmarks
`coo_slice` results in at least two shapes. Should be more useful with the argument `ldk` and within a `Coo`.
```{r}
shp %>% coo_center %>% coo_slice(ids=c(93, 516)) %>% Opn %>% panel()
```

### Indexing
#### Sliding
```{r}
shp %>% coo_plot(main="original cat", cex.first.point = 2)
shp %>% coo_slide(id=93) %>% coo_plot(main="coo_slide", cex.first.point = 2) # same comment for ldk/Coo as for coo_slice
shp %>% coo_slidedirection("right") %>%  coo_plot(main="coo_slidedirection", cex.first.point = 2)
```

#### Closing/opening
```{r}
shp12 <- shp %>% coo_sample(12)
shp12 %>% coo_plot(main="a cubist cat")
shp12 %>% coo_close %>% coo_plot(main="coo_close")
shp12 %>% coo_close %>% coo_unclose %>% coo_plot(main="coo_unclose")
```

#### Trimming/extracting
```{r}
shp12 %>% coo_trimtop(3) %>% coo_plot(main="coo_trimtop")
shp12 %>% coo_trimbottom(3) %>% coo_plot(main="coo_trimtop")
shp12 %>% coo_trim(3) %>% coo_plot(main="coo_trim")
shp %>% coo_extract(93:516) %>% coo_plot(main="coo_extract")
```

#### Reversing
```{r}
shp12 %>% coo_rev %>% coo_plot(main="coo_rev", cex.first.point = 2)
```

### Misc
A side effect of `coo_dxy` is to set the first point on `(0, 0)`

```{r}
shp %>% coo_dxy %>% coo_plot(main="coo_dxy", cex.first.point = 2)
```

`coo_force2close` can be use to spread differences between the first and last point, and to force closing:
```{r}
shp %>% coo_center %>% coo_up %>% coo_plot(main="half a cat")
shp %>% coo_center %>% coo_up %>% coo_force2close %>% coo_plot(main="stop animal cruelty now")
```

```{r}
# for the sake of reproducibility
set.seed(123)
shp %>% coo_jitter(factor=10) %>% coo_plot(main="coo_jitter")
```

## Other coo operations
```{r, eval=FALSE}
# coo_descriptors along
coo_centdist
coo_perimcum
coo_perimpts



# coo_descriptors non-scalars
coo_boundingbox
coo_centpos # should be a df?
coo_chull
coo_chull_onion
coo_diffrange # coo_range_diff
coo_lw
coo_truss

coo_range
coo_range_enlarge

# coo_drawers
coo_arrows
coo_draw
coo_draw_rads
coo_listpanel
coo_lolli
coo_oscillo # deprecate for a proper oscillo
coo_plot
coo_ruban

# coo_others
coo_angle_edges
coo_angle_tangent

coo_intersect_angle
coo_intersect_direction
coo_intersect_segment

# coo_testers
coo_is_closed
coo_likely_anticlockwise
coo_likely_clockwise

# helpers
coo_check
coo_nb

coo_ldk
```
